`{peacesciencer}`: An R Package for Quantitative Peace Science Research

### 2021-12-22 user-guide/ms.Rmd
batch update of stuff

---
title: "`{peacesciencer}`: An R Package for Quantitative Peace Science Research"
thanks: "Replication files are available on the author's Github account (https://github.com/svmiller/peacesciencer). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: steven.v.miller@gmail.com."
author:
- name: Steven V. Miller
  affiliation: Clemson University
bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
# bibliography: inst/refs.bib
abstract: "This article introduces `{peacesciencer}`, an R package that contains a litany of tools for creating data of widespread interest to the peace science community. The package is cross-platform, assuming only a somewhat recent installation of the R programming language with some of the enhanced functionality of the broadly popular `{tidvyerse}` packages. Peace science researchers can use this package to greatly reduce the time needed to perfectly recreate common types of data from scratch and to merge in ubiquitous indicators included in almost every analysis (e.g. democracy data, contiguity data). The software is freely available on CRAN and maintains an active website documenting its features at `r params$ps_website`."
keywords: "software, statistical analysis, peace science, reproducibility"
geometry: margin=1in
csl: inst/jpr.csl
mainfont: cochineal
#sansfont: Linux Biolinum O 
fontsize: 11pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: ""
  removetitleabstract: ""
  ps_website: ""
  ps_manuals: ""
  ps_github: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
#always_allow_html: TRUE
header-includes:
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{csquotes}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
#  - \let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
#  - \usepackage{colortbl}
#  - \arrayrulecolor{black}
#  - \usepackage{threeparttable}
#  - \usepackage{threeparttablex}
#  - \midrule\relax
---

@@

tribble(~func, ~type, ~description,
        "add_archigos()", "D,S", "Add Archigos political leader information to a data frame",
        "add_atop_alliance()", "D, LD", "Add Alliance Treaty Obligations and Provisions (ATOP) alliance data to a data frame",
        "add_capital_distance()", "D, L, LD, S", "Add capital-to-capital distance to a data frame",
        "add_ccode_to_gw()", "D, L, LD, S", "Match CoW state codes to G-W state codes",
        "add_contiguity()", "D, L, LD, S", "Add CoW direct contiguity information to a data frame",
        "add_cow_alliance()", "D, LD", "Add CoW alliance data to a data frame",
        "add_cow_majors()", "D, L, LD, S", "Add CoW major power information to a data frame",
        "add_cow_mids()", "D", "Add CoW-MID data to a dyad-year data frame",
        "add_cow_trade()", "D, L, LD, S", "Add CoW trade data to a data frame",
        "add_cow_wars()", "D, S", "Add CoW war data to a dyad-year or state-year data frame",
        "add_creg_fractionalization()", "D, L, LD, S", "Add fractionalization/polarization estimates from CREG to a data frame",
        "add_democracy()", "D, L, LD, S", "Add democracy information to a data frame",
        "add_fpsim()", "D, LD", "Add dyadic foreign policy similarity measures to a data frame",
        "add_gml_mids()", "D, L, LD, S", "Add GML MID data to a data frame",
        "add_gwcode_to_cow()", "D, L, LD, S", "Match G-W state codes to CoW state codes",
        "add_igos()", "D, S", "Add CoW IGO data to a data frame",
        "add_lead()", "L, LD", "Add leader experience/attributes (LEAD) to data",
        "add_lwuf()", "L, LD", "Add leader willingness to use force estimates to data",
        "add_minimum_distance()", "D, L, LD, S", "Add minimum distance data to data frame",
        "add_nmc()", "D, L, LD, S", "Add minimum distance estimates (in km) to data frame",
        "add_peace_years()", "D, S", "Add 'peace years' to dyad-year/state-year conflict data",
        "add_rugged_terrain()", "D, L, LD, S", "Add rugged terrain information to a data frame",
        "add_sdp_gdp()", "D, L, LD, S", "Add (surplus, gross) domestic product data to a data frame",
        "add_spells()", "D, L, LD, S", "Add duration 'spells' to a data frame",
        "add_strategic_rivalries()", "D, S", "Add strategic rivalry information to a data frame",
        "add_ucdp_acd()", "S", "Add UCDP Armed Conflict Data to state-year data frame",
        "add_ucdp_onsets", "S", "Add UCDP onsets to state-year data",
        "filter_prd()", "D, LD", "Filter dyad-year data to just politically relevant dyads",
        "wc_duration()", "C", "Whittle Duplicate Conflict-Years by (Minimum or Maximum) Conflict Duration",
        "wc_fatality()", "C", "Whittle Duplicate Conflict-Years by Highest Fatality",
        "wc_hostility()", "C", "Whittle Duplicate Conflict-Years by Highest Hostility",
        "wc_jds()", "C", "Whittle Duplicate Conflict-Years by Just Dropping Something",
        "wc_onsets()", "C", "Whittle Uniqute Conflict Onset-Years from Conflict-Year Data",
        "wc_recip()", "C", "Whittle Duplicate Conflict-Years by Conflict Reciprocation",
        "wc_stmon()", "C", "Whittle Duplicate Conflict-Years by Lowest Start Month"
) -> add_functions

@@

```{r tab-functionality, echo=F, eval=F}
tribble(~data, ~description, ~type, ~func,
        "(CoW, G-W) Dyad-Year Data", "Create dyad-years from state system membership data", " ", "create_dyadyears()",
        "(CoW, G-W) State-Year Data", "Create state-years from state system membership data", " ", "create_stateyears()",
        "(CoW, G-W) State-Day Data", "Create state-days from state system membership data", " ", "create_statedays()",
        "Archigos Leader-Day Data", "Create leader-day data from Archigos", " ", "create_leaderdays()",
        "Archigos Leader-Year Data", "Create leader-year data from Archigos", " ", "create_leadeyears()",
        "Archigos Leader-Dyad-Year Data", "Create leader-dyad-year data from Archigos", " ", "create_leaderdyadyears()",
        "Archigos", "Add Archigos political leader information", "D, S", "add_archigos()",
        "ATOP Alliances (v. 5.0)", "Add ATOP alliance data", "D", "add_atop_alliance()",
        "Capital-to-Capital Distances", "Add capital-to-capital distance", "D, S", "add_capital_distance()",
        "Convert CoW codes to G-W codes", "Add matching CoW codes for G-W states by year", "D, S", "add_ccode_to_gw()",
        "Convert G-W codes to CoW codes", "Add matching G-W codes for CoW states by year", "D, S", "add_gwcode_to_cow()",
        "CoW National Capabilities", "Adds capabilities of states by year", "D, S", "add_nmc()",
        "CoW Direct Contiguity", "Adds contiguity information by year", "D, S", "add_contiguity()",
        "CoW Alliances", "Adds CoW alliance data by year", "D", "add_cow_alliance()",
        "CoW IGOs", "Adds CoW IGO data by year", "D, S", "add_igos()",
        "CoW Major Powers", "Adds CoW major power by year", "D, S", "add_cow_majors()",
        "CoW MIDs", "Add CoW MID information to data", "D", "add_cow_mids()",
        "CoW Trade Data", "Add CoW trade information to data", "D, S", "add_cow_trade()",
        "CoW Wars", "Add CoW inter-state (v. 4.0) and intra-state (v. 4.1) war data", "D", "add_cow_wars()",
        "CREG", "Add ethnic/religious fractionalization/polarization by year", "D, S", "add_creg_fractionalization()",
        "Democracy", "Add Polity (v. 2017), V-Dem (v. 10), and Marquez' (2016) UDS extensions", "D,S", "add_democracy()",
        "GML MIDs", "Add Gibler-Miller-Little MIDs to data", "D", "add_gml_mids()",
        "Minimum Distance", "Adds minimum distance data by year", "D, S", "add_minimum_distance()",
        "Peace Years", "Calculate peace years for various disputes", "D, S", "add_peace_years()",
        "Rugged/Mountainous Terrain", "Add 'ruggedness'/'mountainous' terrain data by year", "D, S", "add_rugged_terrain()",
        "(Surplus/Gross) Domestic Product", "Add Anders et al. (2020) GDP/SDP simulations by year", "D, S", "add_sdp_gdp()",
        "Strategic Rivalries", "Add Thompson and Dreyer (2012) strategic rivalry data by year", "D, S", "add_strategic_rivalries()",
        "UCDP Armed Conflicts (v. 20.1)", "Add UCDP armed conflict data by year", "S", "add_ucdp_acd()"
        ) -> included

@@

EUGene is the clear inspiration for this package. Though its original impetus was the generation of expected utility data for evaluating @bdmlalman1992wr (i.e. the "EU" in "EUGene"), the software became quite popular for peace science scholars in the early 2000s for helping them start new projects from scratch with important data already provided. `{peacesciencer}` covers all the same units of analysis that EUGene covers, as of version 3.2.[^v3212] EUGene has more explicit support for dyadic dispute-year data whereas `{peacesciencer}` treats dyadic dispute-year data as a derivation of dyad-year data, with respect to functions that populate base data with additional information.[^whittledeclare] `{peacesciencer}` may reflect more current research interests in the peace science community. This is why there are leader-level data and support for the G-W (and UCDP) ecosystem of data that EUGene does not have, but there is no function yet for things like calculating Tau-b or expected utility values. Importantly, though, `{peacesciencer}` mimics the verbosity of EUGene's user manual. EUGene's user manual was amply informative about what it was calculating and why it was doing what it was doing for some defaults. Likewise, `{peacesciencer}`'s documentation strives to be as informative as possible as to what it is doing and why it is doing what it is doing for some defaults. A scholar who remembers EUGene well will ideally think of `{peacesciencer}` as the most faithful approximation of what that software did for the community at the time. It has the added benefit of being agnostic to the researcher's operating system, having greater flexibility of data types supported, and  better reflecting more current frontiers in the community. It does have the drawback of requiring at least some level of comfort with the R programming language.

@@

Finally, a skeptical reader should not think that making the process as simple as possible necessarily facilitates poor decision-making by the user. In cases where it is evident what the user wants (e.g. an estimate of the level of democracy in the state-year), `{peacesciencer}` does the necessary work to provide the user that information. However, the package makes sure to leave important decision-making to the researcher. For example, `add_cow_alliance()` returns information about various types of alliance pledges in the dyad-year---should one exist---but leaves it to the researcher to say whether they want to define the presence of an alliance to be just a defense pledge or any type of alliance pledge. `add_contiguity()` returns information about the type of contiguity relationship in the dyad-year, but leaves it to the researcher whether they want to code a contiguity variable as the presence of a mutual land border or some other type of contiguity threshold. The documentation included in this package, and on the website, is replete with caveats about the underlying data (e.g. the contiguity data are not ordinal and should not be treated as such), how and where data issues arise (e.g. how CoW state system data differ from Gleditsch-Ward data and how one is coerced into the other), and how researchers should consider optimally using its functionality (e.g. `add_ucdp_acd()` probably should not lump all forms of conflict together). `{peacesciencer}` does not endeavor to make researchers lazy or sloppy, and it does not ultimately do this. Instead, `{peacesciencer}` encourages reasoned design decisions by the user up front and reduces the tedium associated with starting quantitative peace science research. It achieves this in a quick, robust, and transparent way.

>>>>

---
title: "`{peacesciencer}`: An R Package for Quantitative Peace Science Research"
thanks: "Replication files are available on the author's Github account (https://github.com/svmiller/peacesciencer). **Current version**: `r format(Sys.time(), '%B %d, %Y')`; **Corresponding author**: steven.v.miller@gmail.com."
author:
- name: Steven V. Miller
  affiliation: Clemson University
bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
# bibliography: inst/refs.bib
abstract: "This article introduces `{peacesciencer}`, an R package that contains a litany of tools for creating data of widespread interest to the peace science community. The package is cross-platform, assuming only a somewhat recent installation of the R programming language with some of the enhanced functionality of the broadly popular `{tidvyerse}` packages. Peace science researchers can use this package to greatly reduce the time needed to perfectly recreate common types of data from scratch and to merge in ubiquitous indicators included in almost every analysis (e.g. democracy data, contiguity data). The software is freely available on CRAN and maintains an active website documenting its features at `r params$ps_website`."
keywords: "software, statistical analysis, peace science, reproducibility"
geometry: margin=1in
csl: inst/jpr.csl
mainfont: cochineal
#sansfont: Linux Biolinum O 
fontsize: 11pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: ""
  removetitleabstract: ""
  ps_website: ""
  ps_manuals: ""
  ps_github: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
nocite: |
  @cartersmith2020fml, @marshalletal2017p, @coppedgeetal2020vdem, @gleditschetal2002ac, @thompsondreyer2012hir, @cowstates2016, @stinnettetal2002cow, @pevehouseetal2020tow, @schvitz2021mis, @palmeretal2021mid5, @singer1987rcwd, @nardulli2012creg, @gledtischward1999rlis, @nunnpuga2012r, @haege2011cc, @pettersson2019ov, @leedsetal2002atop, @barbierietal2009td
#always_allow_html: TRUE
header-includes:
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{csquotes}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
#  - \let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
#  - \usepackage{colortbl}
#  - \arrayrulecolor{black}
#  - \usepackage{threeparttable}
#  - \usepackage{threeparttablex}
#  - \midrule\relax
---

@@

tribble(~func, ~type, ~description,
        "add_archigos()", "D,S", "Add Archigos political leader information to a data frame",
        "add_atop_alliance()", "D, LD", "Add Alliance Treaty Obligations and Provisions (ATOP) alliance data to a data frame",
        "add_capital_distance()", "D, L, LD, S", "Add capital-to-capital distance to a data frame",
        "add_ccode_to_gw()", "D, L, LD, S", "Match CoW state codes to G-W state codes",
        "add_contiguity()", "D, L, LD, S", "Add CoW direct contiguity information to a data frame",
        "add_cow_alliance()", "D, LD", "Add CoW alliance data to a data frame",
        "add_cow_majors()", "D, L, LD, S", "Add CoW major power information to a data frame",
        "add_cow_mids()", "D", "Add CoW-MID data to a dyad-year data frame",
        "add_cow_trade()", "D, L, LD, S", "Add CoW trade data to a data frame",
        "add_cow_wars()", "D, S", "Add CoW war data to a dyad-year or state-year data frame",
        "add_creg_fractionalization()", "D, L, LD, S", "Add fractionalization/polarization estimates from CREG to a data frame",
        "add_democracy()", "D, L, LD, S", "Add democracy information to a data frame",
        "add_fpsim()", "D, LD", "Add dyadic foreign policy similarity measures to a data frame",
        "add_gml_mids()", "D, L, LD, S", "Add GML MID data to a data frame",
        "add_gwcode_to_cow()", "D, L, LD, S", "Match G-W state codes to CoW state codes",
        "add_igos()", "D, S", "Add CoW IGO data to a data frame",
        "add_lead()", "L, LD", "Add leader experience/attributes (LEAD) to data",
        "add_lwuf()", "L, LD", "Add leader willingness to use force estimates to data",
        "add_minimum_distance()", "D, L, LD, S", "Add minimum distance data to data frame",
        "add_nmc()", "D, L, LD, S", "Add Correlates of War National Military Capabilities data",
        "add_peace_years()", "D, S", "Add 'peace years' to dyad-year/state-year conflict data",
        "add_rugged_terrain()", "D, L, LD, S", "Add rugged terrain information to a data frame",
        "add_sdp_gdp()", "D, L, LD, S", "Add (surplus, gross) domestic product data to a data frame",
        "add_spells()", "D, L, LD, S", "Add duration 'spells' to a data frame",
        "add_strategic_rivalries()", "D, S", "Add strategic rivalry information to a data frame",
        "add_ucdp_acd()", "S", "Add UCDP Armed Conflict Data to state-year data frame",
        "add_ucdp_onsets", "S", "Add UCDP onsets to state-year data",
        "filter_prd()", "D, LD", "Filter dyad-year data to just politically relevant dyads",
        "wc_duration()", "C", "Whittle Duplicate Conflict-Years by (Minimum or Maximum) Conflict Duration",
        "wc_fatality()", "C", "Whittle Duplicate Conflict-Years by Highest Fatality",
        "wc_hostility()", "C", "Whittle Duplicate Conflict-Years by Highest Hostility",
        "wc_jds()", "C", "Whittle Duplicate Conflict-Years by Just Dropping Something",
        "wc_onsets()", "C", "Whittle Uniqute Conflict Onset-Years from Conflict-Year Data",
        "wc_recip()", "C", "Whittle Duplicate Conflict-Years by Conflict Reciprocation",
        "wc_stmon()", "C", "Whittle Duplicate Conflict-Years by Lowest Start Month"
) -> add_functions

@@

```{r tab-functionality, echo=F, eval=F}
tribble(~data, ~description, ~type, ~func,
        "(CoW, G-W) Dyad-Year Data", "Create dyad-years from state system membership data", " ", "create_dyadyears()",
        "(CoW, G-W) State-Year Data", "Create state-years from state system membership data", " ", "create_stateyears()",
        "(CoW, G-W) State-Day Data", "Create state-days from state system membership data", " ", "create_statedays()",
        "Archigos Leader-Day Data", "Create leader-day data from Archigos", " ", "create_leaderdays()",
        "Archigos Leader-Year Data", "Create leader-year data from Archigos", " ", "create_leadeyears()",
        "Archigos Leader-Dyad-Year Data", "Create leader-dyad-year data from Archigos", " ", "create_leaderdyadyears()",
        "Archigos", "Add Archigos political leader information", "D, S", "add_archigos()",
        "ATOP Alliances", "Add ATOP alliance data", "D", "add_atop_alliance()",
        "Capital-to-Capital Distances", "Add capital-to-capital distance", "D, S", "add_capital_distance()",
        "Convert CoW codes to G-W codes", "Add matching CoW codes for G-W states by year", "D, S", "add_ccode_to_gw()",
        "Convert G-W codes to CoW codes", "Add matching G-W codes for CoW states by year", "D, S", "add_gwcode_to_cow()",
        "CoW National Capabilities", "Adds capabilities of states by year", "D, S", "add_nmc()",
        "CoW Direct Contiguity", "Adds contiguity information by year", "D, S", "add_contiguity()",
        "CoW Alliances", "Adds CoW alliance data by year", "D", "add_cow_alliance()",
        "CoW IGOs", "Adds CoW IGO data by year", "D, S", "add_igos()",
        "CoW Major Powers", "Adds CoW major power by year", "D, S", "add_cow_majors()",
        "CoW MIDs", "Add CoW MID information to data", "D", "add_cow_mids()",
        "CoW Trade Data", "Add CoW trade information to data", "D, S", "add_cow_trade()",
        "CoW Wars", "Add CoW inter-state and intra-state war data", "D", "add_cow_wars()",
        "CREG", "Add ethnic/religious fractionalization/polarization by year", "D, S", "add_creg_fractionalization()",
        "Democracy", "Add Polity, V-Dem, and Marquez' (2016) UDS extensions", "D,S", "add_democracy()",
        "GML MIDs", "Add Gibler-Miller-Little MIDs to data", "D", "add_gml_mids()",
        "Minimum Distance", "Adds minimum distance data by year", "D, S", "add_minimum_distance()",
        "Peace Years", "Calculate peace years for various disputes", "D, S", "add_peace_years()",
        "Rugged/Mountainous Terrain", "Add 'ruggedness'/'mountainous' terrain data by year", "D, S", "add_rugged_terrain()",
        "(Surplus/Gross) Domestic Product", "Add Anders et al. (2020) GDP/SDP simulations by year", "D, S", "add_sdp_gdp()",
        "Strategic Rivalries", "Add Thompson and Dreyer (2012) strategic rivalry data by year", "D, S", "add_strategic_rivalries()",
        "UCDP Armed Conflicts", "Add UCDP armed conflict data by year", "S", "add_ucdp_acd()"
        ) -> included

@@

EUGene is the clear inspiration for this package. Though its original impetus was the generation of expected utility data for evaluating @bdmlalman1992wr (i.e. the "EU" in "EUGene"), the software became quite popular for peace science scholars in the early 2000s for helping them start new projects from scratch with important data already provided. `{peacesciencer}` covers all the same units of analysis that EUGene covers, as of version 3.2.[^v3212] EUGene has more explicit support for dyadic dispute-year data whereas `{peacesciencer}` treats dyadic dispute-year data as a derivation of dyad-year data, with respect to functions that populate base data with additional information.[^whittledeclare] `{peacesciencer}` may reflect more current research interests in the peace science community. This is why there are leader-level data and support for the G-W (and UCDP) ecosystem of data that EUGene does not have, but there is no function yet for things like calculating expected utility values. Importantly, though, `{peacesciencer}` mimics the verbosity of EUGene's user manual. EUGene's user manual was amply informative about what it was calculating and why it was doing what it was doing for some defaults. Likewise, `{peacesciencer}`'s documentation strives to be as informative as possible as to what it is doing and why it is doing what it is doing for some defaults. A scholar who remembers EUGene well will ideally think of `{peacesciencer}` as the most faithful approximation of what that software did for the community at the time. It has the added benefit of being agnostic to the researcher's operating system, having greater flexibility of data types supported, and  better reflecting more current frontiers in the community. It does have the drawback of requiring at least some level of comfort with the R programming language.

@@

Finally, a skeptical reader should not think that making the process as simple as possible necessarily facilitates poor decision-making by the user. In cases where it is evident what the user wants (e.g. an estimate of the level of democracy in the state-year), `{peacesciencer}` does the necessary work to provide the user that information. However, the package makes sure to leave important decision-making to the researcher. For example, `add_cow_alliance()` returns information about various types of alliance pledges in the dyad-year---should one exist---but leaves it to the researcher to say whether they want to define the presence of an alliance to be just a defense pledge or any type of alliance pledge. `add_contiguity()` returns information about the type of contiguity relationship in the dyad-year, but leaves it to the researcher whether they want to code a contiguity variable as the presence of a mutual land border or some other type of contiguity threshold. The documentation included in this package, and on the website, is replete with caveats about the underlying data (e.g. the contiguity data are not ordinal and should not be treated as such), how and where data issues arise (e.g. how CoW state system data differ from Gleditsch-Ward data and how one is coerced into the other), and how researchers should consider optimally using its functionality (e.g. `add_ucdp_acd()` probably should not lump all forms of conflict together [c.f. @giblermiller2021apm]). `{peacesciencer}` does not endeavor to make researchers lazy or sloppy, and it does not ultimately do this. Instead, `{peacesciencer}` encourages well-reasoned design decisions by the user up front and reduces the tedium associated with starting quantitative peace science research. It achieves this in a quick, robust, and transparent way.
